dads manual

### 2023-03-27 inst/gitbook/index.Rmd
polished index

--- 
title: "dads manual"
author: "Thomas Guillerme (guillert@tcd.ie), Alex Slavenko (email) and others"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [../references.bib]
link-citations: yes
github-repo: TGuillerme/dads
description: "dads R package vignette"
---

@@

![Schematised summary of the `dads` package architecture](dads_structure.pdf)

@@

 0. **Starting the process**: this step is actually non modular and creates a random tree with one tip, one node and one branch length connecting both. This step is used to optimise the rest of the algorithm in terms of speed and memory management. The node, tip and branch resulting from this step are discarded at the end of the simulations.
 1. **Selecting a lineage** (`selecting`): this step selects a tip that is currently not extinct. In a standard birth-death process this is just done randomly, however in `dads` this can be modified based on the birth-death parameters, the currently available lineages and potential trait values. *For example, it is possible to put a higher probability in selecting a lineage that is next to a lineage that recently went extinct and has a positive trait value.*
 Then go to step 2.
 2. **Growing the tree** (`waiting`): this step grows the tree by a certain amount. It does so by adding the same amount of branch length to all the non-extinct lineages. In an exact birth death process, this is done by drawing a random value from an exponential distribution with a rate of $$\text{number of living lineages} \times (\text{speciation} + \text{extinction parameters})$$. In `dads`, this can be modified based on the birth-death parameters, the currently available lineages and potential trait values. *For example, it is possible to increase branch length by the ratio of living/extinct fossils and a random number drawn from the range of current trait values.*
 If the total tree length is less than the required tree length, go to step 3. Else go to step 6.
 3. **Simulating traits** \[optional\] (`traits`): this step allows to simulate a trait value for the selected lineage from step 1. This is typically not part of a standard birth-death process and is handled via the `traits` option in `dads` (see the [`make.traits`](#maketraits) chapter for more details).
 Then go to step 4.
 4. **Speciating** (`speciating`): in this step, the selected lineage from step 1 has the option of speciating or going extinct. In a standard birth-death process, this happens by randomly drawing a value between 0 and 1 as a way to trigger speciation relatively to the birth-death parameters:
```
if
    randomly drawn number is smaller or equal to speciation/(speciation + extinction)
then
    do speciate
else
    go extinct
```
 Again, in `dads`, this process can be modified based on the birth-death parameters, the currently available lineages and potential trait values. *For example, the lineage can only speciate if it's trait value is positive, regardless of whether speciation or extinction has been triggered*. 
 Then go to step 5.
 5. If the number of lineages is less than required number of lineages, then go to step 1. Else go to step 6.
 6. The simulation has stopped because it reached the required amount of lineages or time.

>>>>

--- 
title: "dads manual"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [../references.bib]
link-citations: yes
github-repo: TGuillerme/dads
description: "dads R package manual"
---

@@

![Schematised summary of the `dads` package architecture](dads_structure.png)

@@

 0. **Starting the process**: this step is actually non modular and creates a random tree with one tip, one node and one branch length connecting both. This step is used to optimise the rest of the algorithm in terms of speed and memory management. The node, tip and branch resulting from this step are discarded at the end of the simulations.
 1. **Selecting a lineage** (`selecting`): this step selects a tip that is currently not extinct. In a standard birth-death process this is just done randomly, however in `dads` this can be modified based on the birth-death parameters, the currently available lineages and potential trait values. *For example, it is possible to put a higher probability in selecting a lineage that is next to a lineage that recently went extinct and has a positive trait value.*
 Then go to step 2.
 2. **Growing the tree** (`waiting`): this step grows the tree by a certain amount. It does so by adding the same amount of branch length to all the non-extinct lineages. In an exact birth death process, this is done by drawing a random value from an exponential distribution with a rate of $$\text{number of living lineages} \times (\text{speciation} + \text{extinction parameters})$$. In `dads`, this can be modified based on the birth-death parameters, the currently available lineages and potential trait values. *For example, it is possible to increase branch length by the ratio of living/extinct fossils and a random number drawn from the range of current trait values.*
 If the total tree length is less than the required tree length, go to step 3. Else go to step 6.
 3. **Simulating traits** [optional]\(`traits`): this step allows to simulate a trait value for the selected lineage from step 1. This is typically not part of a standard birth-death process and is handled via the `traits` option in `dads` (see the [`make.traits`](#maketraits) chapter for more details).
 Then go to step 4.
 4. **Speciating** (`speciating`): in this step, the selected lineage from step 1 has the option of speciating or going extinct. In a standard birth-death process, this happens by randomly drawing a value between 0 and 1 as a way to trigger speciation relatively to the birth-death parameters:
```
if
    randomly drawn number is smaller or equal to speciation/(speciation + extinction)
then
    do speciate
else
    go extinct
```
 Again, in `dads`, this process can be modified based on the birth-death parameters, the currently available lineages and potential trait values. *For example, the lineage can only speciate if it's trait value is positive, regardless of whether speciation or extinction has been triggered*. 
 Then go to step 5.
 5. If the number of lineages is less than required number of lineages, then go to step 1. Else go to step 6.
 6. The simulation has stopped because it reached the required amount of lineages or time.
